Possible Inconsistency in Maximum Likelihood Calculation of Ancestral States

### 2017-07-08 joint_inf.tex
Checking a few things in Felsenstein lower bounds.

Our approach seeks bounds for the terms inside the maximization in \eqref{eq:log_likelihood_simplified}.
We obtain
%EM Does it really make sense to talk about taking the max over \fullAncestralSplitPartitions(\tau,t)? This is already defined. I suggest just taking the max over \fullAncestralSplitPartitions as a variable.
$$
\max_{t} \ \ell_{\tau^*,t^*}(\tau, t; \fullAncestralSplitPartitions(\tau,t)) \le
    \shannonDivergence_{\tau^*,t^*}(\tau^*,t^*)
    + \tilde{\ell}_{\tau^*,t^*}(\tau, \hat{t}; \fullAncestralSplitPartitions(\tau, \hat{t}))
$$
as an upper bound for the joint maximum of \eqref{eq:log_likelihood_simplified} using Gibbs's inequality
$$
\shannonDivergence_{\tau^*,t^*}(\tau,t) \le \shannonDivergence_{\tau^*,t^*}(\tau^*,t^*)
$$
and
$$
\hat{t} = \argmax_{t} \ \tilde{\ell}_{\tau^*,t^*}(\tau, t; \fullAncestralSplitPartitions(\tau,t)).
$$
Similarly, we have the lower bound
$$
\max_{t} \ \ell_{\tau^*,t^*}(\tau, t; \fullAncestralSplitPartitions(\tau,t)) \ge
    \shannonDivergence_{\tau^*,t^*}(\tau,\hat{t})
    + \tilde{\ell}_{\tau^*,t^*}(\tau, \hat{t}; \fullAncestralSplitPartitions(\tau, \hat{t})).
$$

@@

We construct a similar lower bound on the Felsenstein tree partial likelihood.
Replace all $(1+w)$ terms with $(1-w)$ and focus just on the partial likelihood.
Then, if $(1+x)(1-y) > (1-x)(1+y)$,
\begin{align*}
    \tilde{\ell}_{\tau_1,\{x,y,y\}}(\tau_2, \{x',y',w'\}; \fullAncestralSplitPartitions(\tau_2,\{x',y',w'\}))
    &\ge      p_{\emptyset}  \cdot\log((1+x')^2   (1+w')(1+y')^2) \\
    &\qquad + p_{1}          \cdot\log((1+x')^2   (1-w')(1+y')(1-y')) \\
    &\qquad + p_{2}          \cdot\log((1+x')^2   (1-w')(1+y')(1-y')) \\
    &\qquad + p_{3}          \cdot\log((1+x')^2   (1-w')(1+y')(1-y')) \\
    &\qquad + p_{12}         \cdot\log((1+x')^2   (1-w')(1+y')^2) \\
    &\qquad + p_{123}        \cdot\log((1+x')^2   (1-w')(1+y')(1-y'))\\
    &\qquad + p_{13}         \cdot\log((1+x')^2   (1-w')(1-y')^2) \\
    &\qquad + p_{23}         \cdot\log((1+x')^2   (1-w')(1-y')^2),
\end{align*}
otherwise
\begin{align*}
    \tilde{\ell}_{\tau_1,\{x,y,y\}}(\tau_2, \{x',y',w'\}; \fullAncestralSplitPartitions(\tau_2,\{x',y',w'\}))
    &\ge      p_{\emptyset}  \cdot\log((1+x')^2    (1+w')(1+y')^2) \\
    &\qquad + p_{1}          \cdot\log((1+x')(1-x')(1-w')(1+y')^2) \\
    &\qquad + p_{2}          \cdot\log((1+x')(1-x')(1-w')(1+y')^2) \\
    &\qquad + p_{3}          \cdot\log((1+x')(1-x')(1-w')(1+y')^2) \\
    &\qquad + p_{12}         \cdot\log((1+x')^2    (1-w')(1+y')^2) \\
    &\qquad + p_{123}        \cdot\log((1+x')(1-x')(1-w')(1+y')^2)\\
    &\qquad + p_{13}         \cdot\log((1-x')^2    (1-w')(1+y')^2) \\
    &\qquad + p_{23}         \cdot\log((1-x')^2    (1-w')(1+y')^2).
\end{align*}
Letting
$$
b = p_{1}+p_{2}+p_{3}+p_{123}
$$
yields
\begin{align*}
    \tilde{\ell}_{\tau_1,\{x,y,y\}}(\tau_2, \{x',y',w'\}; \fullAncestralSplitPartitions(\tau_2,\{x',y',w'\}))
    &\ge      2\cdot\log(1+x') \\
    &\qquad + (2-b)  \cdot\log(1+y') \\
    &\qquad + b      \cdot\log(1-y') \\
    &\qquad + p_{\emptyset}\cdot\log(1+w') \\
    &\qquad + (1-p_{\emptyset})\cdot\log(1-w')
\end{align*}
for the first lower bound, but, upon maximizing, we obtain
\begin{align*}
    \tilde{\ell}_{\tau_1,\{x,y,y\}}(\tau_2, \{x',y',w'\}; \fullAncestralSplitPartitions(\tau_2,\{x',y',w'\}))
    &\ge      2\cdot\log(2) \\
    &\qquad + (2-b)  \cdot\log(2-b) \\
    &\qquad + b      \cdot\log(b) \\
    &\qquad + p_{\emptyset}\cdot\log(2p_{\emptyset}) \\
    &\qquad + (1-p_{\emptyset})\cdot\log(2(1-p_{\emptyset}))
\end{align*}
maximized at $\{1,1-b,1-2p_{\emptyset}\}$, which is the same if $(1+x)(1-y) > (1-x)(1+y)$ or $(1+x)(1-y) < (1-x)(1+y)$.
Our lower bound is
\begin{align*}
    L_{\tau_1,\tau_2}(\mathbf{p}_{xy}) &:= \shannonDivergence_{\tau_1,\{x,y,y\}}(\tau_2,\{1,1-b,1-2p_{\emptyset}\}) \\
                                           &+ 2\cdot\log(2) + (2-b)  \cdot\log(2-b) + b      \cdot\log(b) + p_{\emptyset}\cdot\log(2p_{\emptyset}) + (1-p_{\emptyset})\cdot\log(2(1-p_{\emptyset})).
\end{align*}

>>>>

Our approach seeks bounds for the terms inside the maximization in \eqref{eq:log_likelihood_simplified}.
We obtain
%EM Does it really make sense to talk about taking the max over \fullAncestralSplitPartitions(\tau,t)? This is already defined. I suggest just taking the max over \fullAncestralSplitPartitions as a variable.
%EM Dave, did you already address this comment? It doesn't seem to make sense anymore.
$$
\max_{t} \ \ell_{\tau^*,t^*}(\tau, t; \fullAncestralSplitPartitions(\tau,t)) \le
    \shannonDivergence_{\tau^*,t^*}(\tau^*,t^*)
    + \tilde{\ell}_{\tau^*,t^*}(\tau, \hat{t}; \fullAncestralSplitPartitions(\tau, \hat{t}))
$$
as an upper bound for the joint maximum of \eqref{eq:log_likelihood_simplified} using Gibbs's inequality
$$
\shannonDivergence_{\tau^*,t^*}(\tau,t) \le \shannonDivergence_{\tau^*,t^*}(\tau^*,t^*)
$$
and
$$
\hat{t} = \argmax_{t} \ \tilde{\ell}_{\tau^*,t^*}(\tau, t; \fullAncestralSplitPartitions(\tau,t)).
$$
Similarly, we have the lower bound
$$
\max_{t} \ \ell_{\tau^*,t^*}(\tau, t; \fullAncestralSplitPartitions(\tau,t)) \ge
    \shannonDivergence_{\tau^*,t^*}(\tau,\hat{t})
    + \tilde{\ell}_{\tau^*,t^*}(\tau, \hat{t}; \fullAncestralSplitPartitions(\tau, \hat{t})).
$$

@@

We construct a similar lower bound on the Felsenstein tree partial likelihood.
In the Felsenstein case, there are various optimal internal states depending on branch lengths (Table~\ref{tab:likelihoods}; $\tau = \tau_2$).
We can bound these below replacing $1+w$ with $1-w$, and by splitting into two cases.
%EM Is this inequality flipped?
If $(1+x)(1-y) > (1-x)(1+y)$,
\begin{align*}
    \tilde{\ell}_{\tau_1,\{x,y,y\}}(\tau_2, \{x',y',w'\}; \fullAncestralSplitPartitions(\tau_2,\{x',y',w'\}))
    &\ge      p_{\emptyset}  \cdot\log((1+x')^2   (1+w')(1+y')^2) \\
    &\qquad + p_{1}          \cdot\log((1+x')^2   (1-w')(1+y')(1-y')) \\
    &\qquad + p_{2}          \cdot\log((1+x')^2   (1-w')(1+y')(1-y')) \\
    &\qquad + p_{3}          \cdot\log((1+x')^2   (1-w')(1+y')(1-y')) \\
%EM it seems to me that the following line should be (1+x')(1-x')(1-w')(1+y')(1-y')
    &\qquad + p_{12}         \cdot\log((1+x')^2   (1-w')(1+y')^2) \\
    &\qquad + p_{123}        \cdot\log((1+x')^2   (1-w')(1+y')(1-y'))\\
    &\qquad + p_{13}         \cdot\log((1+x')^2   (1-w')(1-y')^2) \\
    &\qquad + p_{23}         \cdot\log((1+x')^2   (1-w')(1-y')^2),
\end{align*}
otherwise
\begin{align*}
    \tilde{\ell}_{\tau_1,\{x,y,y\}}(\tau_2, \{x',y',w'\}; \fullAncestralSplitPartitions(\tau_2,\{x',y',w'\}))
    &\ge      p_{\emptyset}  \cdot\log((1+x')^2    (1+w')(1+y')^2) \\
    &\qquad + p_{1}          \cdot\log((1+x')(1-x')(1-w')(1+y')^2) \\
    &\qquad + p_{2}          \cdot\log((1+x')(1-x')(1-w')(1+y')^2) \\
    &\qquad + p_{3}          \cdot\log((1+x')(1-x')(1-w')(1+y')^2) \\
%EM it seems to me that the following line should be (1+x')(1-x')(1-w')(1+y')(1-y')
    &\qquad + p_{12}         \cdot\log((1+x')^2    (1-w')(1+y')^2) \\
    &\qquad + p_{123}        \cdot\log((1+x')(1-x')(1-w')(1+y')^2)\\
    &\qquad + p_{13}         \cdot\log((1-x')^2    (1-w')(1+y')^2) \\
    &\qquad + p_{23}         \cdot\log((1-x')^2    (1-w')(1+y')^2).
\end{align*}
%EM check below after the above is adjusted.
Letting
$$
b = p_{1}+p_{2}+p_{3}+p_{123}
$$
yields
\begin{align*}
    \tilde{\ell}_{\tau_1,\{x,y,y\}}(\tau_2, \{x',y',w'\}; \fullAncestralSplitPartitions(\tau_2,\{x',y',w'\}))
    &\ge      2\cdot\log(1+x') \\
    &\qquad + (2-b)  \cdot\log(1+y') \\
    &\qquad + b      \cdot\log(1-y') \\
    &\qquad + p_{\emptyset}\cdot\log(1+w') \\
    &\qquad + (1-p_{\emptyset})\cdot\log(1-w')
\end{align*}
for the first lower bound, but, upon maximizing, we obtain
\begin{align*}
    \tilde{\ell}_{\tau_1,\{x,y,y\}}(\tau_2, \{x',y',w'\}; \fullAncestralSplitPartitions(\tau_2,\{x',y',w'\}))
    &\ge      2\cdot\log(2) \\
    &\qquad + (2-b)  \cdot\log(2-b) \\
    &\qquad + b      \cdot\log(b) \\
    &\qquad + p_{\emptyset}\cdot\log(2p_{\emptyset}) \\
    &\qquad + (1-p_{\emptyset})\cdot\log(2(1-p_{\emptyset}))
\end{align*}
maximized at $\{1,1-b,1-2p_{\emptyset}\}$, which is the same if $(1+x)(1-y) > (1-x)(1+y)$ or $(1+x)(1-y) < (1-x)(1+y)$.
Our lower bound is
\begin{align*}
    L_{\tau_1,\tau_2}(\mathbf{p}_{xy}) &:= \shannonDivergence_{\tau_1,\{x,y,y\}}(\tau_2,\{1,1-b,1-2p_{\emptyset}\}) \\
                                           &+ 2\cdot\log(2) + (2-b)  \cdot\log(2-b) + b      \cdot\log(b) + p_{\emptyset}\cdot\log(2p_{\emptyset}) + (1-p_{\emptyset})\cdot\log(2(1-p_{\emptyset})).
\end{align*}
